Clustering Beat-Chroma Patterns\\ in a Large Music Database

### 2010-06-01 Paper/Paper.tex
minor
minor tweaks.  made paper fit in 6 pages

What are the common features of pop music?  There are conventions
of subject matter, instrumentation, form, rhythm, harmony, and melody, among
others.  Our interest here is in the tonal content of the music -- \ie the
harmony and melody.  As a computationally-convenient proxy for a richer
description of the tonal content of audio, we use the popular Chroma
representation, which collapses an acoustic spectrum into a 12-dimensional
description, with one bin for each semitone of the western musical octave.
In addition, we simplify the time axis of our representation to take advantage
of the strong beat present in most pop music, and record just one
chroma vector per beat.  This beat-synchronous chromagram
representation represents a typical music track
in a few thousand values, yet when resynthesized back into audio via
modulation of octave-invariant ``Shepard tones'', the melody and
chord sequences of the original music usually remain recognizable
\cite{Ellis2007a}.  To the extent, then, that beat-chroma
representations preserve tonal content, they
are an interesting domain in which to search for patterns -- rich enough
to generate musically-relevant results, but simplified enough to
abstract away aspects of the original audio such as
instrumentation and other stylistic details.

@@

Finally, we normalize the patches with respect to transposition by rotating
the pattern matrix so that the first row contains the most
energy. This can be seen in the example codeword of Figure \ref{fig:code}.
% We considered the same technique for rolling along the other axis and
% be invariant regarding the downbeat. However, it seems less robust to
% noise.
Each patch within a song is normalized independently, so
reconstruction of the original song requires knowledge of the
rotation index for each patch. This rotation removes the key information
of a patch.

@@

For testing, we made use of low quality (32kbps, 8 kHz bandwidth mono MP3) versions of the songs
from the {\it uspop2002} data set \cite{uspop2002}.  This well-known data set 
contains pop songs from a range of artists and styles.
% \footnote{We also tried to use Tzanetakis dataset,
% %\cite{Tzanetakis2002a},
% but the $30$ seconds segments seemed hard to analyze by the Echo Nest API, 
% most did not contain any bar.}.
{\it uspop2002} serves as test set to measure how well a codebook learned on
the Cowbell data set can represent new songs.  We obtained Echo Nest features
for the $8651$ songs contained in the dataset.

\subsection{Settings}\label{ssec:setting}
We take one or two bars and resample the patches to 4 or 8 columns 
respectively.  We learn a codebook of size $K$ over the Cowbell
dataset using the online VQ algorithm (Algorithm \ref{algo:vq}). We use a
learning rate of $\ell=0.01$ for $200$ iterations over the whole
dataset.
%
We then use the resulting codebook to encode the test set.
Each pattern is encoded with only one code. We can measure the average
distance between a pattern and its encoding. We can also measure the use
of the different codes, 
%\ie the size of the cluster around a code.
i.e., the proportion of patterns that quantize to each code.

@@

We trained a codebook $200$ entries of patterns sized $12\times8$, covering
$2$ bars at a time.  The results shown are on the {\it artist20} test set described in 
Section \ref{ssec:artist}.

@@

\section{Applications}\label{sec:exps2}
We present two applications of the beat-chroma codebooks
to illustrate how the ``natural'' structure identified via 
unsupervised clustering can provide useful 
features for subsequent supervised tasks.  We will discuss how 
the codewords can be used in bar alignment, and for artist
recognition.  
%The representation created from large collections
%captures meaningful information without being forced to do so.
%
% Although our algorithm was not devised for any of these
% tasks, but some information contained in the centroids can be
% naturally used.

@@

\begin{figure}[h]
  \centering
  \subfloat[][\small{Metallica}]{\label{fig:metallica2}\includegraphics[width=0.32\columnwidth]{metallica_pattern}}
  \hspace{5mm}                
  \subfloat[][\small{Tori Amos / Suzanne Vega}]{\label{fig:amosvega2}\includegraphics[width=0.32\columnwidth]{toriamos_suzannevega_pattern}}
  \vspace{2mm}
  \caption{\small{Typical patterns for different artists.}}
  \label{fig:typicalpat}
\end{figure}

@@

%\small
\section{Acknowledgements}
Thanks to Graham Grindlay for numerous discussions and helpful comments.
T. Bertin-Mahieux is supported by a NSERC scholarship.
This material is based upon work supported by the IMLS (grant LG-06-08-0073-08)
and by NSF grant IIS-0713334. Any opinions, findings and
conclusions or recommendations expressed in this material are those of
the authors and do not necessarily reflect the views of the sponsors

>>>>

What are the common features of pop music?  There are conventions
of subject matter, instrumentation, form, rhythm, harmony, and melody, among
others.  Our interest here is in the tonal content of the music -- \ie the
harmony and melody.  As a computationally-convenient proxy for a richer
description of the tonal content of audio, we use the popular chroma
representation, which collapses an acoustic spectrum into a 12-dimensional
description, with one bin for each semitone of the western musical octave.
In addition, we simplify the time axis of our representation to take advantage
of the strong beat present in most pop music, and record just one
chroma vector per beat.  This beat-synchronous chromagram
representation represents a typical music track
in a few thousand values, yet when resynthesized back into audio via
modulation of octave-invariant ``Shepard tones'', the melody and
chord sequences of the original music usually remain recognizable
\cite{Ellis2007a}.  To the extent, then, that beat-chroma
representations preserve tonal content, they
are an interesting domain in which to search for patterns -- rich enough
to generate musically-relevant results, but simplified enough to
abstract away aspects of the original audio such as
instrumentation and other stylistic details.

@@

Finally, we normalize the patches with respect to transposition by rotating
the pattern matrix so that the first row contains the most
energy. This can be seen in the example codeword of Figure \ref{fig:code}.
% We considered the same technique for rolling along the other axis and
% be invariant regarding the downbeat. However, it seems less robust to
% noise.
Each patch within a song is normalized independently, so
reconstruction of the original song requires knowledge of the
rotation index for each patch.
%This rotation removes the key information of a patch.

@@

For testing, we made use of low quality (32kbps, 8 kHz bandwidth mono MP3) versions of the songs
from the {\it uspop2002} data set \cite{uspop2002}.  This data set 
contains pop songs from a range of artists and styles.
% \footnote{We also tried to use Tzanetakis dataset,
% %\cite{Tzanetakis2002a},
% but the $30$ seconds segments seemed hard to analyze by the Echo Nest API, 
% most did not contain any bar.}.
{\it uspop2002} serves as test set to measure how well a codebook learned on
the Cowbell data set can represent new songs.  We obtained Echo Nest features
for the $8651$ songs contained in the dataset.

\subsection{Settings}\label{ssec:setting}
We take one or two bars and resample the patches to 4 or 8 columns 
respectively.  We learn a codebook of size $K$ over the Cowbell
dataset using the online VQ algorithm (Algorithm \ref{algo:vq}). We use a
learning rate of $\ell=0.01$ for $200$ iterations over the whole
dataset.
%
We then use the resulting codebook to encode the test set.
Each pattern is encoded with a single code. We can measure the average
distance between a pattern and its encoding. We can also measure the use
of the different codes, 
%\ie the size of the cluster around a code.
i.e., the proportion of patterns that quantize to each code.

@@

We trained a codebook containing $200$ patterns sized $12\times8$, covering
$2$ bars at a time.  The results shown are on the {\it artist20} test set described in 
Section \ref{ssec:artist}.

@@

\section{Applications}\label{sec:exps2}
We present two applications of the beat-chroma codebooks
to illustrate how the ``natural'' structure identified via 
unsupervised clustering can provide useful 
features for subsequent supervised tasks.  We will discuss how 
the codewords can be used in bar alignment, and artist
recognition.  
%The representation created from large collections
%captures meaningful information without being forced to do so.
%
% Although our algorithm was not devised for any of these
% tasks, but some information contained in the centroids can be
% naturally used.

@@

\begin{figure}[t]
  \centering
  \subfloat[][\small{Metallica}]{\label{fig:metallica2}\includegraphics[width=0.32\columnwidth]{metallica_pattern}}
  \hspace{5mm}                
  \subfloat[][\small{Tori Amos / Suzanne Vega}]{\label{fig:amosvega2}\includegraphics[width=0.32\columnwidth]{toriamos_suzannevega_pattern}}
  \vspace{2mm}
  \caption{\small{Typical patterns for different artists.}}
  \label{fig:typicalpat}
\end{figure}

@@

%\small
\section{Acknowledgements}
Thanks to Graham Grindlay for numerous discussions and helpful comments.
T. Bertin-Mahieux is supported by a NSERC scholarship.
This material is based upon work supported by IMLS grant LG-06-08-0073-08
and by NSF grant IIS-0713334. Any opinions, findings and
conclusions or recommendations expressed in this material are those of
the authors and do not necessarily reflect the views of the sponsors.
